---
title: "Example application of the Cologne Protocol for the Early Neolithic of central Europe"
author: "M. Broich & R. Peters"
date: "11 March 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This manual presents an example application of the **Cologne Protocol** published in *Schmidt et al. (accepted)*. It is based upon the `.R` files which are in the `code\` directory. If you want to apply the Cologne Protocol on your own data, we recommend you to use these files, as it is possible to individually adjust certain variables at the beginning of the `00_LEC.R` file. 

The manual describes the first two parts of the Cologne Protocol, which are a GIS-analysis of site distribution and the identifying of so-called Core Areas. The aim of this document is neither to explain the further steps of the Cologne Protocol nor the theoretical background. For these points please refer to the associated publication including the supplementary information.

The goal of the script is to carry out all steps of the first part of the Cologne Protocol in R. These are (cf. *Schmidt et al. accepted, table S2*):

1. **Calculating "Largest Empty Circle"**

    i) Creating of Voronoi Polygons
    ii) Extraction of Vertices
    iii) Aggregation of Verticies
    iv) Defining the Radius of the "Largest Empty Circle"

2. **Kriging**

    i) Preparation and Grid
    ii) Semivariogram
    iii) Inspect and Raster Output

3. **Optimally Describing Isolines**

    i) Creating Contour Lines (Isolines)
    ii) Calculating the Area and Number of Sites per Ssoline
    iii) Selecting the "Optimally Describing Isoline"

4. **Data export**


## Dependencies

To run the code of the script several packages are needed. These dependencies are stored in the `deps.yaml` file. It is possible to load them manually or to use `automagic::install_deps_file()` function. If the package `automagic` is not installed, it is possible to intsall it with `install.packages("automagic")`.  
Furthermore the script was developed under R version 3.6.3.


## Sample Data

For this examplary application, we use a distribution map of the Early Neolithic Linear Pottery Culture (LBK) in Central Europe. The map is based on *Preuss (1998, Karte 1)* and available from the CRC 806 database (https://crc806db.uni-koeln.de/start/).

The distribution map can be loaded into R via URL: 

```{r load data}
# url_link of distribution map of Linear Pottery Culture
url_link <- "http://sfb806srv.uni-koeln.de/owsproxy.php?service=WFS&version=1.0.0&request=GetFeature&typeNames=geonode%3A_13_earlyneolithic_ce_sites_wgs84&outputFormat=csv"

# load date as a data.frame
sites <- read.csv(url(url_link))

# Conversion into SPDF
sites <- sp::SpatialPointsDataFrame(sp::SpatialPoints(
  cbind(sites$RECHTS, sites$HOCH)),
  sites,
  proj4string = CRS("+init=epsg:25832"))

# This is not a reprojection, but some function if the projection is not defined like this
sp::proj4string(sites) <- sp::CRS("+init=epsg:25832")

```

It is important to note, that cartesian coordinates (planar coord.) have to be used, because `rgeos::gDistance` will only accept planar coordinates. Especially Lat/Long-coordinates will cause a problem. 

## Largest Empty Circle

### Creating of Voronoi polygons

### Extraction of vertices

### Aggregation of verticies

### Defining the radius of the "Largest Empty Circle"

## Kriging

### Preparation and Grid

### Semivariogram

### Inspect and Raster Output

## Optimally Describing Isolines

### Creating Contour Lines (Isolines)

### Calculating the Area and Number of Sites per Isoline

### Selecting the "Optimally Describing Isoline"

## Data Export

## Bibliography

Preuss, J. (ed.) **1998**. Das Neolithikum in Mitteleuropa: Kulturen, Wirtschaft, Umwelt vom 6. bis 3. Jahrtausend v. u. Z., Übersichten zum Stand der Forschung (Weissbach 1998).

Schmidt, I., Hilpert, J., Kretschmer, I., Peters, R., Broich, M., Schiesberg, S., Vogels, O., Wendt, K. P., Zimmermann, A., Maier, A., **accepted**. Approaching Prehistoric Demography: Proxies, Scales and Scope of the Cologne Protocol in European contexts. _Philosophical Transactions B_.
